Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-07-09 article.tex
changes in phrasing, update intro fig

\documentclass[pdflatex]{sn-jnl}
\jyear{2024}
\usepackage{multibib}
\newcites{m}{Methods References}
\usepackage[superscript]{cite}
\usepackage{caption}
\bibliographystyle{unsrt}
\bibliographystylem{unsrt} 
\raggedbottom

@@

\abstract{
Earthquake occurrence is notoriously difficult to predict. While some aspects of their spatiotemporal statistics can be relatively well captured by point-process models, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts. This is due both to the lack of information about fault conditions and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter distribution, which describes the marginal distribution over large regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

@@

This modeling approach is not a naive or uninformed choice.  Current research on magnitude prediction has failed to show a universal and reproducible advantage over a stationary or quasi-stationary GR benchmark~\cite{ogata_exploring_2018, stockman_forecasting_2023}. Indeed, due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event requires a detailed microscopic knowledge of the system's state. 
This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}. These models reproduce many of the phenomenological statistical relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event is indeed impossible using only far-field measurements.

Multiple previous studies have not identified a definitive link between earthquake magnitude and seismic history \cite{petrillo_verifying_2023, taroni_are_2024, davidsen_are_2011}. Some studies have shown circumstantial evidence that earthquake magnitudes may not follow a strictly stationary distribution, e.g.~by showing that statistics are not invariant to permutation of the order, or by showing that the maximal magnitude in a given time frame can be predicted with a better-than-random chance~\cite{xiong_seismic_2023, corral_comment_2005, spassiani_exploring_2016, lippiello_positive_2024, lippiello_positive_2024-1, lippiello_influence_2008, shcherbakov_forecasting_2019, panakkat_neural_2007}. However, a significant information gain over the GR benchmark in predicting the magnitude of a given earthquake was not demonstrated, and to the best of our knowledge was not attempted. The debate regarding the dependence of magnitude on seismic history remains open.

@@

% \neri{Stress more: this will imply we have information relevant to prediction tasks. Zhuang stated: "you have touched one of the most important problems on earthquake predictability"}

To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence. Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude. In some experiments, we also incorporate focal mechanism information. Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided. Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.

% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

\section{Methodology}
We construct the MAGnitude Neural EsTimation model, MAGNET, a generative neural network (NN) that receives as input a hypocentral catalog of past regional seismicity, and the time and location of a future event (a ``query''). The network produces a probability density function (PDF) estimating its magnitude.

MAGNET is composed of two main components. First, the catalog up to time $t$ (not including) is encoded by long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig. \ref{fig:intro_fig}a.

@@

We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate} for Japan. While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the PDF. The resulting PDF is presented in Fig. \ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark.

\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/intro_fig.png}
	\caption{\textbf{a} MAGnitude Neural EsTimation model (MAGNET) task and architecture. Seismic history is continuously encoded using a Long Short-Term Memory neural network. For each earthquake, the encoded history up to its occurrence time ($t_i$) is concatenated with the time and location of the earthquake (``spacetime query''). The combined information is fed into a fully-connected neural network that outputs a probability density function (PDF) of the predicted magnitude at that time and location. \textbf{b} Performance on major earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide. The red curve represents the model's predicted PDF and the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark. For most earthquakes shown here, the likelihood of the observed magnitude (red star) is higher for MAGNET than for the GR benchmark (black circle). Notably, the Tōhoku aftershock shows a qualitatively different behavior.
    }
\label{fig:intro_fig}
\end{figure}

All metrics reported in this work were evaluated over a time span that is not included in the training period. The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating a significant information gain in forecasting earthquake magnitudes prior to their occurrence.
%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation. Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.

@@

This is a qualitative demonstration that the model works as expected. For a quantitative comparison, we calculate the $\mathcal{L}=-\langle \ell_i \rangle$ over all events in the test set, and benchmark it against state-of-the-art models:
the stationary Gutenberg-Richter (GR) distribution (fitted over the training set), and a few variants of a moving window GR approach~\cite{gulia_real-time_2019}. It is seen that MAGNET significantly and consistently outperforms all benchmarks across all three regions, as shown in Fig. \ref{fig:metrics}a-c. Further comparison to other benchmarks is given in Table \ref{tab:mean_ll_benchmarks_main_text}.

@@

Finally, Figure~\ref{fig:model_output}b presents the marginal probability density function (PDF) of magnitudes for MAGNET's prediction, $P(m)=N^{-1}\sum_i p_{\textbf{x}_i, t_i}(m)$. This represents the average probability distribution of magnitudes over the whole test set. It is seen that although the model was not constrained to do so, the average distribution aligns well with the ``uninformed'' GR distribution. That is, although MAGNET's prediction for each individual earthquake may deviate significantly from the GR distribution, it does reproduce this well known long-time behavior on average. It is also seen that the average prediction of the 300-event moving window GR deviates significantly from the GR distribution. We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data. This observation is noticeable even more so in the New Zealand and Japan data sets presented in Extended Data Fig. \ref{fig:model_output_em}.

@@

\section{Interpretation and caveats}
At this point, it is clear that MAGNET is able to extract information about the magnitude of specific earthquakes, prior to their occurrence. However, such an information gain may potentially result not from the fact that the underlying physical process is predictable (at least to some extent), but from various measurement and modeling artifacts.

@@

Similarly to the temporal domain, a spurious information gain may also be caused by temporally constant but spatially varying properties. For example, fault stresses, geometry and lithology may change in space resulting in a spatially-dependent GR distribution \cite{amitrano_brittle-ductile_2003, scholz_stress_2015, herrmann_revealing_2022, taroni_earthquake_2023}, and the local completeness magnitude is also generally spatially dependent. In such a case, the information gain of MAGNET does not imply that the magnitudes are correlated in time. 
To examine this, we preform two additional tests. First, we construct a spatially dependent GR distributions where $m_c$ is held constant ($=m_c^{\mbox{\footnotesize (train set)}}$). MAGNET outperforms the spatially varying GR benchmark for all regions in the unconditioned $\mathcal{L}$ score, as can be seen in EM Table \ref{tab:mean_ll_all_benchmarks}. The spatially varying GR shows an advantage over MAGNET for solely the New Zealand data set for the temporally conditioned $\mathcal{L}$. This advantage may arise from a the measurement artifact of spatial variation of $m_c$ originating from the properties of the seismic network.

In order to account for such spatially dependent artifacts we preform a simialr calculation. We once again condition MAGNET's results on a threshold according to Eq. \ref{eq:conditioned_likelihood}, this time choosing a spatially varying completeness $\Tilde{m}=m_c(x,y)$. We estimate $m_c(x,y)$ for a grid of coordinates spaced every $0.1^\circ$, for all events within a square centered at each grid point, with a side length of $0.2^\circ$.

The spatially conditioned CIG curves are shown as the red dashed curve in Fig. \ref{fig:info_gain_over_time} for all three regions. It is a steadily increasing function for the Southern California and Japan data sets. In the New Zealand data set the curve shows an initial increase followed by a plateau after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes). This may indicate that a good portion of this information gain followed by these earthquakes originate in the spatial incompleteness artifact. Nevertheless, the overall score still shows an advantage for our model over the test set. Fig. \ref{fig:info_gain_over_time}e-g show the spatial distribution of MAGNET's advantage over the GR benchmark. All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area.

\section{Implications and Outlook}
In this work we have used neural based models to find statistical dependencies between earthquake magnitudes and seismic history. Since our model is provided with the known location and timing of an event and is only tasked with predicting its magnitude, we have separated out the question of rate and location of earthquakes, and directly probe whether their magnitude is history dependent. MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent incompleteness, compared against a host of benchmarks. Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a simple hypo-center catalog.

%A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated.

If true, this means that the seismic patterns preceding large events are distinguishable, at least to some extent, from those preceding ``background'' activity. A similar claim was recently raised by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}, which show trends in statistical properties of the seismicity prior to fault-size events, though a direct linkage between the two observation requires further investigation.

In the future, it will also be interesting to examine MAGNET's prediction on query coordinates in the vicinity (either in space or time) of true events, producing a space-time hazard map, which may be used to inform existing forecasting and early-warning systems. This can also be done by constructing a different model, trained to estimate the magnitude of an upcoming earthquake ahead of the origin time, examining how it depends on the lead time and distance from the event location. 
It would also be interesting to understand what are the statistical features of seismicity that our model identifies, which may be done using various explainability analyses~\cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023}. Potentially, ablation tests \cite{meyes_ablation_2019, sturmfels_visualizing_2020} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.

\section{Possible effects on operational earthquake forecasting and prediction}
Identifying precursory patterns in the seismic history can advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}. Nevertheless, discovering such signals with enough accuracy to prove useful for operational systems may still be further down the road.
However, an immediate application of the model may be a quick estimation of magnitude for early warning systems. Such a system does not require additional data sources or assumptions on the model as the timing and location of the earthquake to be estimated are known or estimated. 
Another immediate application is to incorporate such a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}. As discussed above, such models currently assume no statistical dependence of the magnitude. Since larger earthquakes produce more aftershock, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.

% \neri{I bet the reviewer will want us to talk about limitations and improvements to the model but maybe we can wait and see}

>>>>

\documentclass[pdflatex]{sn-jnl}
\jyear{2024}
\usepackage{multibib}
\newcites{m}{Methods References}
\usepackage[superscript]{cite}
\usepackage{caption}
\usepackage[left]{lineno}
\linenumbers
\bibliographystyle{unsrt}
\bibliographystylem{unsrt} 
\raggedbottom

@@

\abstract{
Earthquake occurrence is notoriously difficult to predict. While some aspects of their spatiotemporal statistics can be relatively well captured by point-process models, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts. This is due both to the lack of information about fault conditions and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over large regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present MAGNET - MAGnitude Neural EsTimation model, an open-source, geophysically-inspired neural-network model for probabilistic forecasting of future magnitudes from cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

@@

This modeling approach is not a naive or uninformed choice.  Current research on magnitude prediction has failed to show a universal and reproducible advantage over a stationary or quasi-stationary GR benchmark~\cite{ogata_exploring_2018, stockman_forecasting_2023}. Indeed, due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event requires a detailed microscopic knowledge of the system's state. 
This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}. These models reproduce many of the phenomenological statistical relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically. The magnitudes then follow a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event is indeed impossible using only far-field measurements.

Multiple previous studies have not identified a definitive link between earthquake magnitude and seismic history \cite{petrillo_verifying_2023, taroni_are_2024, davidsen_are_2011}. Some studies have shown circumstantial evidence that earthquake magnitudes may not follow a strictly stationary distribution, e.g.~by showing that statistics are not invariant to permutation of the order, or by showing that the maximal magnitude in a given time frame can be predicted with a better-than-random chance~\cite{xiong_seismic_2023, corral_comment_2005, spassiani_exploring_2016, lippiello_positive_2024, lippiello_positive_2024-1, lippiello_influence_2008, shcherbakov_forecasting_2019, panakkat_neural_2007}. However, a significant information gain over the GR benchmark in predicting the magnitude of a given earthquake was not demonstrated. The debate regarding the dependence of magnitude on seismic history remains open.

@@

To this end, we construct MAGnitude Neural EsTimation model, \nobreak MAGNET, a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence. Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude. In some experiments, we also incorporate focal mechanism information. Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided. Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.

\section{Methodology}
MAGNET is a generative neural network (NN) that takes as input a hypocentral catalog of regional seismicity, and the time and location of a future event (a ``query''). The network produces a probability density function (PDF) estimating its magnitude.

MAGNET is composed of two main components. First, the catalog up to time $t$ (excluding $t$) is encoded by a series of geophysically inspired functions and long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig. \ref{fig:intro_fig}a.

@@

We applied MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate} for Japan. 
While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies and exhibit varying data quality. 
A separate model is trained for each region, with identical loss function and parameterization of the PDF. All metrics reported in this work were evaluated over a time span that is not included in the training period.

The resulting PDF is presented in Fig. \ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark. 
It can be seen that for most presented earthquakes, the likelihood of the observed magnitude is higher for MAGNET than for the GR benchmark, and in some cases MAGNET's prediction differs qualitatively as well as quantitatively.
The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating a significant information gain in forecasting earthquake magnitudes prior to their occurrence.

\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/intro_fig.png}
	\caption{\textbf{a} MAGnitude Neural EsTimation model (MAGNET) task and architecture. Seismic history is continuously encoded using a Long Short-Term Memory neural network. For each earthquake, the encoded history up to its occurrence time ($t_i$) is concatenated with the time and location of the earthquake (``space~-~time query''). The combined information is fed into a fully-connected neural network that outputs a probability density function (PDF) of the predicted magnitude at that time and location. In turn, the output is compared to a benchmark model (e.g. GR in dashed line) \textbf{b} Performance on major earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide. The red curve represents the model's predicted PDF and the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark. For most earthquakes shown here, the likelihood of the observed magnitude (red star) is higher for MAGNET than for the GR benchmark (black circle). Notably, the Tōhoku aftershock shows a qualitatively different behavior.
    }
\label{fig:intro_fig}
\end{figure}

@@

This is a qualitative demonstration that the model works as expected. For a quantitative comparison, we calculate $\mathcal{L}=-\langle \ell_i \rangle$ over all events in the test set, and benchmark it against state-of-the-art models:
the stationary Gutenberg-Richter (GR) distribution (fitted over the training set), and a few variants of a moving window GR approach~\cite{gulia_real-time_2019}. It is seen that MAGNET significantly and consistently outperforms all benchmarks across all three regions, as shown in Fig. \ref{fig:metrics}a-c. Further comparison to other benchmarks is given in Table \ref{tab:mean_ll_benchmarks_main_text}.

@@

Finally, Figure~\ref{fig:model_output}b presents the marginal probability density function (PDF) of magnitudes for MAGNET's prediction, $P(m)=N^{-1}\sum_i p_{\textbf{x}_i, t_i}(m)$. This represents the average probability distribution of magnitudes over the whole test set. It is seen that although the model was not constrained to do so, the average distribution aligns well with the ``uninformed'' GR distribution. That is, although MAGNET's prediction for each individual earthquake may deviate significantly from the GR distribution, it does reproduce this well known long-time behavior on average. It is also seen that the average prediction of the 300-event moving window GR deviates significantly from the GR distribution. We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data. This observation is more pronounced in the New Zealand and Japan data sets presented in Extended Data Fig. \ref{fig:model_output_em}.

@@

\section{Interpretation and potential caveats}
At this point, it is clear that MAGNET achieves an information gain about the magnitude of specific earthquakes, prior to their occurrence. However, such a gain may potentially result not from the fact that the underlying physical process is predictable (at least to some extent), but from various measurement and modeling artifacts.

@@

Similarly to the temporal domain, a spurious information gain may also be caused by temporally constant but spatially varying statistical properties. For example, fault stresses, geometry and lithology may change in space resulting in a spatially-dependent GR distribution \cite{amitrano_brittle-ductile_2003, scholz_stress_2015, herrmann_revealing_2022, taroni_earthquake_2023}, and the local completeness magnitude is also generally spatially dependent. In such a case, the information gain of MAGNET does not imply that the magnitudes are correlated in time.

To examine this, we preform two additional tests.
First, we construct a spatially dependent GR distribution which accounts for the local varation in seismic statistics. We use the method proposed by \cite{taroni_highdefinition_2021}, as detailed in the Methods section. The spatially dependent GR shows some information gain over plain-vanilla GR in New Zealand and Japan, but not in California, see EM Table \ref{tab:mean_ll_all_benchmarks}. In any case, this information gain is always smaller than MAGNET's, with the sole exception of the temporally-conditioned test in New Zealand.

Second, simliar to the temporal analysis, we calculate the information gain conditioned on the locally-varying completness magnitude. Specifically, we use Eq. \ref{eq:conditioned_likelihood}, with a spatially varying completeness $\Tilde{m}=m_c(x,y)$, estimated for a coordinate grid of $0.1^\circ$, see Methods section for details.

The spatially conditioned CIG are shown by the red dashed curves in Fig. \ref{fig:info_gain_over_time}. It is a steadily increasing function for the Southern California and Japan data sets. In the New Zealand data set the curve shows an initial increase followed by a plateau after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes). This may indicate that a good portion of this information gain followed by these earthquakes does originate from spatially inhomogeneity, as also hinted by the spatially-varying GR benchmark. Nevertheless, the overall score still shows an advantage for our model over the test set.
Fig. \ref{fig:info_gain_over_time}e-g shows the spatial distribution of MAGNET's advantage over the GR benchmark. All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area.

\section{Implications and Outlook}
In this work we have used neural based models to find statistical dependencies between earthquake magnitudes and seismic history. Since our model is provided with the known location and timing of an event and is only tasked with predicting its magnitude, we have separated out the question of rate and location of earthquakes, and directly probe whether their magnitude is history dependent. MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent properties compared against a host of benchmarks, thereby answering an open question in earthquake predictability. Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a simple hypo-center catalog.

This means that the seismic patterns preceding large events are distinguishable, at least to some extent, from those preceding ``background'' activity. A similar claim was recently raised by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}, which show trends in statistical properties of the seismicity prior to fault-size events, though a direct linkage between the two observation requires further investigation. 
In general, it would also be interesting to understand what are the statistical features of seismicity that our model identifies, which may be done using various explainability analyses~\cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023}.

Understanding such patterns will advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}, though it is still further down the road.
However, even as a ``black box'', such a model may be integrated into operational systems such early warning systems, which already provide a quick estimation of the source time and location. 
Another possible application is incorporating a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}. As discussed above, such models currently assume no statistical dependence of the magnitude. Since larger earthquakes produce more aftershocks, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.

% In light of the foregoing, our study demonstrates the possibility of extracting information about the magnitude of an upcoming earthquake from seismic history, thereby answering an open question in earthquake predictability. Our work has been put to test against multiple benchmark models and across multiple regions worldwide with various seismogenic properties, demonstrating our conclusions as general. Our model is openly available. To this end, our model has the potential not only to be used for scientific inquiry, but also to be put to use in operational means. We believe it may further advance our understanding and monitoring of earthquakes.
